---
title: "002-covid_therapeutics_USA"
output:
html_document:
df_print: paged
---

<br><br><br>

#### Load packages
```{r}
library(tidyverse)
library(readxl)
library(dplyr)
library(rlist)
```

<br><br><br>

#### Data sources

<br>

therapeuticsData from 'https://healthdata.gov/Health/COVID-19-Public-Therapeutic-Locator/rxn6-qnx8/'

citypopulationData (for cities)  from 'https://www2.census.gov/programs-surveys/popest/datasets/2010-2020/cities/'

therapeuticsDistributionData from 'https://www.phe.gov/emergency/events/COVID19/therapeutics/distribution/Pages/data-tables.aspx'

statePopData from 'https://www2.census.gov/programs-surveys/popest/datasets/2020-2021/state/totals/'

stateCasesData from 'https://data.cdc.gov/Case-Surveillance/United-States-COVID-19-Cases-and-Deaths-by-State-o/9mfq-cb36/data'

<br><br><br>

#### Load in therapeutics data, combine, add date fields, tidy up
```{r}
therapeuticsFileNames <- list.files("01-dataCollection/02-availableCoursesData", pattern="*.csv", full.names=TRUE)

therapeuticsData <- lapply(therapeuticsFileNames, function(x) {
  
  rowDate <- substr(str_split(x, "_", n = Inf, simplify = FALSE)[[1]][[2]], 1, 10)  
  
  dailyData <- read.csv(x) |>
    mutate(rowDate = rowDate)

}) |>
  bind_rows() |>
  mutate(rowYear = format(as.Date(rowDate, format="%Y-%m-%d"),"%Y")) |>
  mutate(monthYear = format(as.Date(rowDate, format="%Y-%m-%d"),"%m")) |>
  mutate(City = str_to_title(City)) |>
  filter(!is.na(Courses.Available)) |>
  mutate(Order.Label = case_when(
    Order.Label == 'Molnupiravir  [Qty 20]' ~ 'Molnupiravir',
    TRUE ~ Order.Label
  ))
```

<br><br><br>

#### Load in other data
```{r}
therapeuticsDistributionData <- read_excel("01-dataCollection/03-distributionData/PHE-Historical data collection.xlsx", sheet = "CombinedData")

cityPopulationData <- read.csv("01-dataCollection/SUB-EST2020_ALL.csv")

statePopData <- read.csv("01-dataCollection/nst-est2020.csv")

stateCasesData <- read.csv("01-dataCollection/01-stateCasesData/United_States_COVID-19_Cases_and_Deaths_by_State_over_Time.csv")
```

<br><br><br>

#### Aggregate 2022 and last seven days Covid-19 case data
```{r}
statesCasesData <- stateCasesData |>
  mutate(rowDate = as.Date(submission_date, format="%m/%d/%Y")) |>
  mutate(dateYear = format(rowDate,"%Y")) |>
  mutate(dateMonth = format(rowDate,"%m")) |>
  mutate(daysFromLastSubmission = as.numeric(str_replace(difftime(max(rowDate), rowDate, units = "days"), " days", "")))

statesCases2022Data <- statesCasesData |>
  filter(dateYear >= 2022) |>
  group_by(state) |>
  summarise(cases2022 = sum(new_case))

statesCasesLastSevenDays <- statesCasesData |>
  filter(daysFromLastSubmission < 7) |>
  group_by(state) |>
  summarise(casesLastSevenDays = sum(new_case))

statesCasesCombined <- merge(x = statesCases2022Data, y = statesCasesLastSevenDays, by = "state", all.x = TRUE)
```

<br><br><br>

#### Tidy up state and city population data
```{r}
statePopData <- statePopData |>
  mutate(stateCode = state.abb[match(NAME,state.name)]) |>
  filter(!is.na(stateCode))

cityPopulationData <- cityPopulationData |>
  mutate(stateCode = state.abb[match(STNAME,state.name)]) |>
  filter(grepl(" city", NAME, fixed = TRUE) & POPESTIMATE2020 > 10000) |>
  mutate(cityName = str_to_title(str_replace(NAME, " city", "")))

cityPopulationData <- cityPopulationData[c('cityName', 'stateCode', 'POPESTIMATE2020')]
cityPopulationData <- cityPopulationData[order(-cityPopulationData$POPESTIMATE2020),]
cityPopulationData <- cityPopulationData[!duplicated(cityPopulationData$cityName, cityPopulationData$stateCode),]
```

<br><br><br>

#### Therapeutics analysis by date (by state and by city)
```{r}
byDateTherapeuticsAnalysis <- lapply(unique(therapeuticsData$rowDate), function(x) {
  
  print(x)
  
  filteredData <- filter(therapeuticsData, rowDate == x)
  
  output <- list()
  
  output[['date']] <- x
  
  statesAnalysis <- filteredData |>
    group_by(Order.Label, State.Code) |>
    summarise(count = sum(Courses.Available)) |>
    pivot_wider(names_from = Order.Label, values_from = count)
    
  statesAnalysis[is.na(statesAnalysis)] <- 0
  
  statesAnalysis <- merge(x = statesAnalysis, y = statesCasesCombined, by.x = "State.Code", by.y = "state", all.x = TRUE)
  
  statesAnalysis <- merge(x = statesAnalysis, y = statePopData[c("stateCode", "POPESTIMATE2020")], by.x = "State.Code", by.y = "stateCode", all.x = TRUE) |>
    filter(!is.na(POPESTIMATE2020))
  
  statesAnalysis <- statesAnalysis |>
    mutate(cases2022Share = cases2022/sum(statesAnalysis$cases2022)) |>
    mutate(cases7DaysShare = casesLastSevenDays/sum(statesAnalysis$casesLastSevenDays)) |>
    mutate(paxlovidShare = Paxlovid/sum(statesAnalysis$Paxlovid)) |>
    mutate(molnupiravirShare = Molnupiravir/sum(statesAnalysis$Molnupiravir)) |>
    mutate(pPlusMShare = (Paxlovid+Molnupiravir)/(sum(statesAnalysis$Paxlovid)+sum(statesAnalysis$Molnupiravir))) |>
    mutate(populationShare = POPESTIMATE2020/sum(statesAnalysis$POPESTIMATE2020)) |>
    mutate(cases2022Per100000 = cases2022*100000/POPESTIMATE2020) |>
    mutate(cases7DaysPer100000 = casesLastSevenDays*100000/POPESTIMATE2020) |>
    mutate(paxlovidCoursesPer100000 = Paxlovid*100000/POPESTIMATE2020) |>
    mutate(molnupiravirCoursesPer100000 = Molnupiravir*100000/POPESTIMATE2020) |>
    mutate(pPlusMCoursesPer100000 = (Paxlovid+Molnupiravir)*100000/POPESTIMATE2020) |>
    mutate(cases7DaysPerPaxlovid = casesLastSevenDays/Paxlovid) |>
    mutate(cases7DaysPerMolnupiravir = casesLastSevenDays/Molnupiravir) |>
    mutate(cases7DaysperPPlusM = casesLastSevenDays/(Paxlovid+Molnupiravir) ) |>
    mutate(rowDate = x)
  
  statesAnalysis <- statesAnalysis[order(-statesAnalysis$pPlusMCoursesPer100000),]
  
  output[['statesAnalysis']] <- statesAnalysis
  
  citiesAnalysis <- filteredData |>
    group_by(Order.Label, State.Code, City) |>
    summarise(count = sum(Courses.Available)) |>
    pivot_wider(names_from = Order.Label, values_from = count)
  
  citiesAnalysis[is.na(citiesAnalysis)] <- 0
  
  citiesAnalysis <- merge(x = citiesAnalysis, y = cityPopulationData, by.x=c('City', 'State.Code'), by.y=c('cityName', 'stateCode'), all.x = TRUE)|>
    mutate(paxlovidShare = Paxlovid/sum(statesAnalysis$Paxlovid)) |>
    mutate(molnupiravirShare = Molnupiravir/sum(statesAnalysis$Molnupiravir)) |>
    mutate(pPlusMShare = (Paxlovid+Molnupiravir)/(sum(statesAnalysis$Paxlovid)+sum(statesAnalysis$Molnupiravir))) |>
    mutate(populationShare = POPESTIMATE2020/sum(statesAnalysis$POPESTIMATE2020)) |>
    mutate(paxlovidCoursesPer100000 = Paxlovid*100000/POPESTIMATE2020) |>
    mutate(molnupiravirCoursesPer100000 = Molnupiravir*100000/POPESTIMATE2020) |>
    mutate(pPlusMCoursesPer100000 = (Paxlovid+Molnupiravir)*100000/POPESTIMATE2020) |>
    mutate(rowDate = x)
  
  citiesAnalysis <- filter(citiesAnalysis, POPESTIMATE2020 > 200000)
  citiesAnalysis <- citiesAnalysis[order(-citiesAnalysis$pPlusMCoursesPer100000),]
  
  output[['citiesAnalysis']] <- citiesAnalysis
  
  return(output)
})
```


<br><br><br>

#### Data extraction (just used for checking which states/figures to include in the copy)
```{r}
dataExtraction1 <- list.filter(byDateTherapeuticsAnalysis, date == '2022-02-09') [[1]][["statesAnalysis"]]

write.table(dataExtraction1, "02-rOutputs/dataExtraction1.csv", row.names = FALSE, col.names = TRUE, sep = ",")
```

<br><br><br>

#### Graph 1 - Covid-19 cases per course - last 7 days (states)
```{r}
graph1Data <- list.filter(byDateTherapeuticsAnalysis, date == '2022-02-09') [[1]][["statesAnalysis"]][c("State.Code", "cases7DaysPerPaxlovid", "cases7DaysPerMolnupiravir", "cases7DaysperPPlusM")]

write.table(graph1Data, "02-rOutputs/graph1Data.csv", row.names = FALSE, col.names = TRUE, sep = ",")
```

<br><br><br>

#### Graph 2 - Available courses per 100,000 (states)
```{r}
graph2Data <- list.filter(byDateTherapeuticsAnalysis, date == '2022-02-09') [[1]][["statesAnalysis"]][c("State.Code", "paxlovidCoursesPer100000", "molnupiravirCoursesPer100000", "pPlusMCoursesPer100000")]

write.table(graph2Data, "02-rOutputs/graph2Data.csv", row.names = FALSE, col.names = TRUE, sep = ",")
```

<br><br><br>

#### Graph 3 - Correlation of courses to cases (states)
```{r}
graph3Data <- list.filter(byDateTherapeuticsAnalysis, date == '2022-02-09') [[1]][["statesAnalysis"]][c("State.Code", "cases7DaysPer100000", "cases2022Per100000", "paxlovidCoursesPer100000", "molnupiravirCoursesPer100000", "pPlusMCoursesPer100000")]

write.table(graph3Data, "02-rOutputs/graph3Data.csv", row.names = FALSE, col.names = TRUE, sep = ",")
```

<br><br><br>

#### Graph 4 - Available course by date (states)
```{r}
graph4Data <- lapply(byDateTherapeuticsAnalysis, function(x) {
  
  output <- x[["statesAnalysis"]][c("State.Code", "paxlovidCoursesPer100000", "rowDate")]
}) |>
  bind_rows() |>
  pivot_wider(names_from = "rowDate", values_from = "paxlovidCoursesPer100000")

write.table(graph4Data, "02-rOutputs/graph4Data.csv", row.names = FALSE, col.names = TRUE, sep = ",")
```

<br><br><br>

#### Graph 5 - Available course by date (states)
```{r}
graph5Data <- lapply(byDateTherapeuticsAnalysis, function(x) {
  
  output <- x[["statesAnalysis"]][c("State.Code", "molnupiravirCoursesPer100000", "rowDate")]
}) |>
  bind_rows() |>
  pivot_wider(names_from = "rowDate", values_from = "molnupiravirCoursesPer100000")

write.table(graph5Data, "02-rOutputs/graph5Data.csv", row.names = FALSE, col.names = TRUE, sep = ",")
```

<br><br><br>

#### Graph 6 - state distribution analysis by date 
```{r}
therapeuticsDistributionData <- therapeuticsDistributionData |>
  mutate(stateCode = state.abb[match(NameGrouped,state.name)]) |>
  filter(!is.na(stateCode)) 

therapeuticsDistributionData <- merge(x = therapeuticsDistributionData, y = statePopData[c("stateCode", "POPESTIMATE2020")], by.x = "stateCode", by.y = "stateCode", all.x = TRUE) |>
  mutate(dosesPer100000 = DoseNumbers*100000/POPESTIMATE2020)

graph6Data <- list.filter(byDateTherapeuticsAnalysis, date == '2022-01-24') [[1]][["statesAnalysis"]][c("State.Code","paxlovidCoursesPer100000", "molnupiravirCoursesPer100000", "pPlusMCoursesPer100000")]

unique(therapeuticsDistributionData$TherapeuticGrouped)

paxlovidDistributionData <- therapeuticsDistributionData |>
  filter(TherapeuticGrouped == 'Paxlovid') |>
  group_by(DateMidpoint, stateCode) |>
  summarise(count = sum(dosesPer100000)) |>
  pivot_wider(names_from = 'DateMidpoint', values_from = 'count')

paxlovidDistributionData <- merge(x = paxlovidDistributionData, y = graph6Data[c("State.Code", "paxlovidCoursesPer100000")], by.x = "stateCode", by.y = "State.Code", all.x = TRUE)

write.table(paxlovidDistributionData, "02-rOutputs/graph6Data-i.csv", row.names = FALSE, col.names = TRUE, sep = ",")

molnupiravirDistributionData <- therapeuticsDistributionData |>
  filter(TherapeuticGrouped == 'Molnupiravir') |>
  group_by(DateMidpoint, stateCode) |>
  summarise(count = sum(dosesPer100000)) |>
  pivot_wider(names_from = 'DateMidpoint', values_from = 'count')

molnupiravirDistributionData <- merge(x = molnupiravirDistributionData, y = graph6Data[c("State.Code", "molnupiravirCoursesPer100000")], by.x = "stateCode", by.y = "State.Code", all.x = TRUE)

write.table(molnupiravirDistributionData, "02-rOutputs/graph6Data-ii.csv", row.names = FALSE, col.names = TRUE, sep = ",")
```

<br><br><br>

#### Get lookup data for population and state name (for graphs)
```{r}
write.table(statePopData[c("NAME", "stateCode", "POPESTIMATE2020")], "02-rOutputs/stateNamePopLookup.csv", row.names = FALSE, col.names = TRUE, sep = ",")
```



